Foregrounds in Wide--Field Redshifted 21~cm Power Spectra

### 2015-02-10 ms.tex
Rename and renumber S5 and its subsections
from 45a74e9:
============

In ms.tex,

(*) Chnage title of S5 to Murchison Widefield Array
(*) Rename and renumber subsections under S5
(*) Change wording in caption of figure on wide-field effects slightly
(*) Change wording in subsection Delay Space slightly
(*) Change wording in caption of Delay space figure slightly
(*) Change wording in subsection Instrument model slightly
(*) Change wording in caption of MWA data vs. model comparison figure
slightly
(*) Change wording in caption of simulated CSM and DSM delay spectra
figure slightly
(*) Change

{\bf We give a brief overview of some parameters of Fourier space which is generic to all experiments that use a similar approach}. Figure~\ref{fig:fourier-space} illustrates the Fourier space in which the delay (and power) spectra of H{\sc i} from the EoR are presented in this paper. $|\boldsymbol{b}|$ and $k_\perp$, denoting spatial scales in the transverse direction {\bf (tangent plane to the celestial sphere)}, form the $x$--axis. $\tau$ and $k_\parallel$, denoting spatial scales along line--of--sight form the $y$--axis. Foreground emission maps to a wedge--shaped region in Fourier space, hereafter referred to as the {\it foreground wedge} \citep{dat10}, whose boundaries are determined by the antenna spacings and the light travel times across them. These boundaries, called horizon delay limits \citep{ved12,par12b}, are shown by solid lines.

\begin{figure}[htb]
\centering
% \includegraphics[width=\linewidth]{figures/v1_0/fourier_space_185.0_MHz_30.7_MHz.eps}
% \includegraphics[width=\linewidth]{f1.eps}
% \includegraphics[width=\linewidth]{fig2.eps}
% \includegraphics[width=\linewidth]{figure2.eps}
\includegraphics[width=\linewidth]{figure1.eps}
\caption{Fourier space in which delay (and power) spectra of EoR H{\sc i} signals are presented. The $x$--axis is denoted by $|\boldsymbol{b}|$ (antenna spacing) or $k_\perp$ (transverse wavenumber). The $y$--axis denoted by $\tau$ (delay) or $k_\parallel$ (line--of--sight wavenumber). Here, $k_\perp$ and $k_\parallel$ are obtained for a frequency of 185~MHz. The dark shaded region is referred to as the {\it foreground wedge} where smooth spectrum foregrounds reside. Its boundaries (solid lines), given by light travel time for corresponding antenna spacings, are referred to as horizon delay limits. Regions excluding the {\it wedge} and narrow extensions of the {\it wedge} (white unshaded strips) caused by convolution with the {\bf instrument's spectral transfer function} are expected to be relatively free of foreground contamination {\bf and are generally referred to as the {\it EoR window}. There are undesirable grating responses (dot--dashed lines) specific to the MWA. Hence, we conservatively identify a restricted region of high EoR sensitivity (medium shade) and refer to it as the MWA {\it EoR window}}. \label{fig:fourier-space}}
\end{figure}

@@

We obtain the delay spectrum of visibilities by taking the delay transform of each baseline's spectrum (equation~\ref{eqn:delay-transform}) choosing $W'_f$ to be a {\it Blackman--Harris} window function. The sky spectrum is multiplied by weights, $W_f$, due to the instrumental passband and frequency channels possibly flagged for radio frequency interference. In delay--space, these weights translate into a convolution by a point spread function (PSF). We deconvolve this PSF using a one dimensional CLEAN algorithm \citep{tay99} as described for the delay axis \citep{par09,par12b} to obtain the final delay spectra. The CLEAN procedure iteratively finds and subtracts peak values convolved by the Fourier transform of the weights. We limit the selection of peaks to modes inside the horizon delay limit, corresponding to smooth spectrum objects in the visible sky hemisphere.

@@

Figure~\ref{fig:usm} shows the wide--field delay power spectrum of a uniformly illuminated sky with no spectral variation measured by antenna elements arranged identical to that in the MWA antenna array layout \citep{bea12} with a uniform power pattern across the sky and a bandwidth of 30.72~MHz centered around 185~MHz (refer to \S\ref{sec:instrument_model} for a detailed description of the instrument model). Notice the steep rise in power towards the horizon limits. These wide--field effects are prevalent on all antenna spacings, including the longest ones used in this study.

@@

{\bf In our present study, we use a latitude of -26.701\arcdeg~ and an antenna layout identical to that of the MWA \citep{bea12} for the observatory. The array is arranged as a centrally condensed core of $\sim$~300~m --- there are many spacings in the range 5--50~m --- and a radial density that falls off as the inverse of the radius, with the longest baselines at 3~km. Here we focus on antenna spacings $|\boldsymbol{b}| \le 200$~m (spatial scales relevant to reionization). Their deviation from coplanarity is negligible. For geometrical intuition, we restrict the orientation ($\theta_b$, measured anti-clockwise from East) of all baselines to lie in the range $-67.5\arcdeg \leq \theta_b < 112.5\arcdeg$. Baselines oriented in the other half--plane measure conjugate visibilities with delays of equal magnitude but of opposite sign and hence are ignored in our analysis. We choose an observing frequency of 185~MHz ($z\simeq 6.68$) and a flat passband of width $\Delta B = 30.72$~MHz to roughly match those of ongoing MWA EoR observations}.

@@

\begin{figure*}[htb]
\centering
% \subfloat[][Power patterns]{\label{fig:power-patterns}\includegraphics[width=\linewidth]{figure5a.eps}} \\
% \subfloat[][delay power spectra]{\label{fig:antennas-delay-spectra}\includegraphics[width=\linewidth]{figure5b.eps}}
\subfloat[][Power patterns]{\label{fig:power-patterns}\includegraphics[width=\linewidth]{figure4a.eps}} \\
\subfloat[][delay power spectra]{\label{fig:antennas-delay-spectra}\includegraphics[width=\linewidth]{figure4b.eps}}
\caption{Power patterns (top panels) and simulated delay power spectra (bottom panels) for different antenna shapes at 185~MHz centered on zenith. Antenna shapes used are: dipole (left), phased array (middle), and dish (right). Refer to text for details. The strength and occupancy of the power patterns are correlated with those of delay power spectra. White dotted lines in the delay power spectra mark the boundaries of the {\it foreground wedge} determined by the horizon delay limit and antenna spacing. The {\it foreground wedge} and the {\it EoR window} are most severely contaminated in the case of the dipole while it is the least for the dish. The phased array has intermediate levels of contamination. Foreground power close to the horizon delay limits in all three cases are significant even on long baselines. The foreshortening of baselines towards the horizon makes them sensitive to foreground emission on large size scales. The amplitude of this feature is highest for the dipole and least for the dish and hence, strongly depends on the shape of the antenna element.}
\label{fig:aperture-shapes}
\end{figure*}

@@

\section{MWA Data Modeling}\label{sec:MWA}

We now use our simulations to analyze features in the delay power spectrum obtained using the MWA.

\subsection{The Murchison Widefield Array (MWA)}\label{sec:instrument}

MWA construction was completed in 2012 and, after commissioning, began its EoR observing program in 2013. In its final configuration, the MWA is a 128--tile interferometer capable of observing a 30.72~MHz instantaneous band anywhere in the range 80--300~MHz. Each tile is a phased array of 16 dipoles, each in the shape of a bow--tie. This yields a primary field of view $\gtrsim$~20\arcdeg~ wide and multiple secondary lobes. The array layout is described in \citet{bea12}. %{\bf At the observing frequency of 185~MHz, we use $A_\textrm{e}=10.5$~m$^2$}.

@@

\subsection{EoR Observations}\label{sec:obsparms}

@@

\begin{figure*}[htb]
\centering
% \subfloat[][Diffuse Sky Model]{\label{fig:DSM}\includegraphics[width=0.45\linewidth]{figure6a.eps}}
% \subfloat[][Bright Point Source Model]{\label{fig:CSM}\includegraphics[width=0.45\linewidth]{figure6b.eps}}
\subfloat[][Diffuse Sky Model]{\label{fig:DSM}\includegraphics[width=0.45\linewidth]{figure5a.eps}}
\subfloat[][Bright Point Source Model]{\label{fig:CSM}\includegraphics[width=0.45\linewidth]{figure5b.eps}}
\caption{Sky brightness temperature of the diffuse foreground model (left) and flux densities of bright point sources (right) at 185~MHz visible during {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The color scales are logarithmic. Power pattern contours are overlaid. The contour levels shown are 0.00195, 0.00781, 0.0312, 0.125, and 0.5. The Galactic center and a portion of the Galactic plane are prominently visible during the {\it off--zenith} pointing in the diffuse sky model and the MWA tile power gain is significant ($\gtrsim 12$\%) in that direction. In contrast, emission from the Galactic plane in {\it zenith} pointing is significantly lesser.}
\label{fig:sky-model-MWA}
\end{figure*}

@@

\subsection{Comparison with Data}\label{sec:data-vs-model}

@@

\begin{figure*}[htb]
\centering
% \includegraphics[width=\linewidth]{figure7.eps}
\includegraphics[width=\linewidth]{figure6.eps}
\caption{Delay power spectra from MWA data (left) and modeling (right) for the {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The {\it foreground wedge} is bounded by white dotted lines. Model matches the data to a level consistent with the uncertainties in foreground models and the antenna beam. \label{fig:fhd-sim-comparison}}
\end{figure*}

@@

\begin{figure*}[htb]
\centering
% \includegraphics[width=\linewidth]{figure8.eps}
\includegraphics[width=\linewidth]{figure7.eps}
\caption{Simulated {\bf delay power spectra (in units of K$^2$~(Mpc/$h$)$^3$} for {\it off--zenith} (top) and {\it zenith} (bottom) pointings for the diffuse (left) and compact (right) foreground models without any thermal noise. The axes and color scale are identical to those in Figure~\ref{fig:fhd-sim-comparison}. In the {\it off-zenith} pointing, emission from the Galactic center is the most prominent feature seen as a branch at $\tau<0$. In the {\it zenith} pointing, {\bf delay power spectrum} from diffuse emission has a {\it two--pronged fork}--shaped structure and is present even at wide antenna spacings due to wide--field effects. Compact emission is centrally concentrated. This combination gives rise to a {\it pitchfork} signature in the {\it foreground wedge}. \label{fig:noiseless-dsm-csm-delay-spectrum}}
\end{figure*}

@@

\begin{figure}[htb]
\centering
% \includegraphics[width=\linewidth]{figure9.eps}
\includegraphics[width=\linewidth]{figure8.eps}
\caption{Simulated {\bf delay power spectra} for three chosen northward oriented antenna spacings of length: $\sim$84~m ({\it top}), $\sim$104~m ({\it middle}), and $\sim$171~m ({\it bottom}). The baseline length and orientation is specified in each panel. The solid red, cyan, and black lines denote contributions from diffuse, compact, and composite foreground models respectively. Vertical dotted lines mark the horizon delay limits. {\bf Compact emission dominates the central regions of the delay power spectra while both components, especially diffuse emission on short antenna spacings, dominate near the horizon delay limits, giving rise to a {\it pitchfork}--shaped structure. The shaded region denotes the envelope of uncertainty around the composite model delay power spectrum. This uncertainty is dictated by uncertainties in spectral indices of foreground models inside the delay horizon delay limits and by thermal noise fluctuations outside}. \label{fig:pitchfork-baselines}}
\end{figure}

>>>>

{\bf We give a brief overview of some parameters of Fourier space which is generic to all experiments that use a similar approach}. Figure~\ref{fig:fourier-space} illustrates the Fourier space in which the delay (and power) spectra of redshifted H{\sc i} observations are calculated. $|\boldsymbol{b}|$ and $k_\perp$, denoting spatial scales in the transverse direction {\bf (tangent plane to the celestial sphere)}, form the $x$--axis. $\tau$ and $k_\parallel$, denoting spatial scales along line--of--sight form the $y$--axis. Foreground emission maps to a wedge--shaped region in Fourier space, hereafter referred to as the {\it foreground wedge} \citep{dat10}, whose boundaries are determined by the antenna spacings and the light travel times across them. These boundaries, called horizon delay limits \citep{ved12,par12b}, are shown by solid lines.

\begin{figure}[htb]
\centering
% \includegraphics[width=\linewidth]{figures/v1_0/fourier_space_185.0_MHz_30.7_MHz.eps}
% \includegraphics[width=\linewidth]{f1.eps}
% \includegraphics[width=\linewidth]{fig2.eps}
% \includegraphics[width=\linewidth]{figure2.eps}
\includegraphics[width=\linewidth]{figure1.eps}
\caption{Fourier space in which delay (and power) spectra of EoR H{\sc i} signals are calculated. The $x$--axis is denoted by $|\boldsymbol{b}|$ (antenna spacing) or $k_\perp$ (transverse wavenumber). The $y$--axis denoted by $\tau$ (delay) or $k_\parallel$ (line--of--sight wavenumber). Here, $k_\perp$ and $k_\parallel$ are obtained for a frequency of 185~MHz. The dark shaded region is referred to as the {\it foreground wedge} where smooth spectrum foregrounds reside. Its boundaries (solid lines), given by light travel time for corresponding antenna spacings, are referred to as horizon delay limits. Narrow extensions of the {\it wedge} (white unshaded strips) are caused by convolution with the {\bf instrument's spectral transfer function}. Regions excluding the {\it wedge} are expected to be relatively free of foreground contamination {\bf and are generally referred to as the {\it EoR window}. There are undesirable grating responses (dot--dashed lines) specific to the MWA. Hence, we conservatively identify a restricted region of high EoR sensitivity (medium shade) and refer to it as the MWA {\it EoR window}}. \label{fig:fourier-space}}
\end{figure}

@@

We obtain the delay spectrum of visibilities by taking the delay transform of each baseline's spectrum (equation~\ref{eqn:delay-transform}) choosing $W'_f$ to be a {\it Blackman--Harris} window function. The sky spectrum is multiplied in the instrument by weights, $W_f$, due to the instrumental passband and flagging of frequency channels possibly corrupted by radio frequency interference, which together are represented by the weights, $W_f$. In delay--space, these weights translate into a convolution by a point spread function (PSF). We deconvolve this PSF using a one dimensional CLEAN algorithm \citep{tay99} as described for the delay axis \citep{par09,par12b} to obtain the final delay spectra. The CLEAN procedure iteratively finds and subtracts peak values convolved by the Fourier transform of the weights. We limit the selection of peaks to modes inside the horizon delay limit, corresponding to smooth spectrum objects in the visible sky hemisphere.

@@

Figure~\ref{fig:usm} shows the wide--field delay power spectrum of a uniformly illuminated sky with no spectral variation as measured by antenna elements arranged identical to that in the MWA antenna array layout \citep{bea12} with a uniform power pattern across the sky and a bandwidth of 30.72~MHz centered around 185~MHz (refer to \S\ref{sec:instrument_model} for a detailed description of the instrument model). Notice the steep rise in power towards the horizon limits. These wide--field effects are prevalent on all antenna spacings, including the longest ones used in this study.

@@

{\bf In our present study, we use a latitude of -26.701\arcdeg~ and an antenna layout identical to that of the MWA \citep{bea12} for the observatory. The array is arranged as a centrally condensed core of $\sim$~300~m --- there are many spacings in the range 5--50~m --- and a radial density that falls off as the inverse of the radius, with the longest baselines at 3~km. Here we focus on antenna spacings $|\boldsymbol{b}| \le 200$~m (spatial scales relevant to reionization). Their deviation from coplanarity is negligible. For geometrical intuition, we restrict the orientation ($\theta_b$, measured anti-clockwise from East) of all baselines to lie in the range $-67.5\arcdeg \leq \theta_b < 112.5\arcdeg$. Baselines oriented in the other half--plane measure conjugate visibilities with delays of equal magnitude but of opposite sign and hence are ignored in our analysis. We choose an observing frequency of 185~MHz ($z\simeq 6.68$) and a flat passband of width $\Delta B = 30.72$~MHz to roughly match those of ongoing MWA EoR observations (discussed in detail in \S\ref{sec:MWA})}.

@@

\begin{figure*}[htb]
\centering
% \subfloat[][Power patterns]{\label{fig:power-patterns}\includegraphics[width=\linewidth]{figure5a.eps}} \\
% \subfloat[][delay power spectra]{\label{fig:antennas-delay-spectra}\includegraphics[width=\linewidth]{figure5b.eps}}
\subfloat[][Power patterns]{\label{fig:power-patterns}\includegraphics[width=\linewidth]{figure4a.eps}} \\
\subfloat[][delay power spectra]{\label{fig:antennas-delay-spectra}\includegraphics[width=\linewidth]{figure4b.eps}}
\caption{Power patterns (top panels) and simulated delay power spectra (bottom panels) for different antenna shapes at 185~MHz centered on zenith. Antenna shapes used are: dipole (left), phased array (middle), and dish (right). Refer to \S\ref{sec:instrument_model} for details of the antenna models. The strength and occupancy of the power patterns are correlated with those of delay power spectra. White dotted lines in the delay power spectra mark the boundaries of the {\it foreground wedge} determined by the horizon delay limit and antenna spacing. The {\it foreground wedge} and the {\it EoR window} are most severely contaminated in the case of the dipole while it is the least for the dish. The phased array has intermediate levels of contamination. Foreground power close to the horizon delay limits in all three cases are significant even on long baselines. The foreshortening of baselines towards the horizon makes them sensitive to foreground emission on large size scales. The amplitude of this feature strongly depends on the shape of the antenna element. It is highest for a dipole (which has a strong response near the horizon) and least for a dish.}
\label{fig:aperture-shapes}
\end{figure*}

@@

\section{The Murchison Widefield Array}\label{sec:MWA}

We now use our simulations to analyze features in the observed delay power spectrum obtained using the MWA.

% \subsection{The Murchison Widefield Array (MWA)}\label{sec:instrument}

MWA construction was completed in 2012 and, after commissioning, began its EoR observing program in 2013. The MWA is a 128--tile interferometer capable of observing a 30.72~MHz instantaneous band anywhere in the range 80--300~MHz. Each tile is a phased array of 16 dipoles, each in the shape of a bow--tie. This yields a primary field of view $\gtrsim$~20\arcdeg~ wide and multiple secondary lobes. The array layout is described in \citet{bea12}. %{\bf At the observing frequency of 185~MHz, we use $A_\textrm{e}=10.5$~m$^2$}.

@@

% \subsection{EoR Observations}\label{sec:obsparms}

@@

\begin{figure*}[htb]
\centering
% \subfloat[][Diffuse Sky Model]{\label{fig:DSM}\includegraphics[width=0.45\linewidth]{figure6a.eps}}
% \subfloat[][Bright Point Source Model]{\label{fig:CSM}\includegraphics[width=0.45\linewidth]{figure6b.eps}}
\subfloat[][Diffuse Sky Model]{\label{fig:DSM}\includegraphics[width=0.45\linewidth]{figure5a.eps}}
\subfloat[][Bright Point Source Model]{\label{fig:CSM}\includegraphics[width=0.45\linewidth]{figure5b.eps}}
\caption{Sky brightness temperature of the diffuse foreground model (left) and flux densities of bright point sources (right) at 185~MHz visible during {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The color scales are logarithmic. MWA tile power pattern contours are overlaid. The contour levels shown are 0.00195, 0.00781, 0.0312, 0.125, and 0.5. The Galactic center and a portion of the Galactic plane are prominently visible during the {\it off--zenith} pointing in the diffuse sky model and the MWA tile power gain is significant ($\gtrsim 12$\%) in that direction. In contrast, emission from the Galactic plane in {\it zenith} pointing is significantly less.}
\label{fig:sky-model-MWA}
\end{figure*}

@@

\subsection{Comparison of Data and Model}\label{sec:data-vs-model}

@@

\begin{figure*}[htb]
\centering
% \includegraphics[width=\linewidth]{figure7.eps}
\includegraphics[width=\linewidth]{figure6.eps}
\caption{Delay power spectra from MWA data (left) and modeling (right) for the {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The {\it foreground wedge} is bounded by white dotted lines. Model matches the data to a level consistent with the uncertainties in foreground models and the antenna beam (as discussed in \S\ref{sec:data-vs-model}). \label{fig:fhd-sim-comparison}}
\end{figure*}

@@

\begin{figure*}[htb]
\centering
% \includegraphics[width=\linewidth]{figure8.eps}
\includegraphics[width=\linewidth]{figure7.eps}
\caption{Simulated {\bf delay power spectra (in units of K$^2$~(Mpc/$h$)$^3$)} for for the diffuse (left) and compact (right) foreground models in the {\it off--zenith} (top) and {\it zenith} (bottom) pointings without any thermal noise. The axes and color scale are identical to those in Figure~\ref{fig:fhd-sim-comparison}. In the {\it off-zenith} pointing, emission from the Galactic center is the most prominent feature seen as a branch at $\tau<0$. In the {\it zenith} pointing, {\bf delay power spectrum} from diffuse emission has a {\it two--pronged fork}--shaped structure and is present even at wide antenna spacings due to wide--field effects. Compact emission is centrally concentrated. \label{fig:noiseless-dsm-csm-delay-spectrum}}
\end{figure*}

@@

\begin{figure}[htb]
\centering
% \includegraphics[width=\linewidth]{figure9.eps}
\includegraphics[width=\linewidth]{figure8.eps}
\caption{Simulated {\bf delay power spectra} for three chosen northward oriented antenna spacings of length: $\sim$84~m ({\it top}), $\sim$104~m ({\it middle}), and $\sim$171~m ({\it bottom}). The baseline length and orientation is specified in each panel. The solid red, cyan, and black lines denote contributions from diffuse, compact, and composite foreground models respectively. Vertical dotted lines mark the horizon delay limits. {\bf Compact emission dominates the central regions of the delay power spectra while both components, especially diffuse emission on short antenna spacings, dominate near the horizon delay limits, giving rise to a characteristic {\it pitchfork}--shaped structure. The shaded region denotes the envelope of uncertainty around the composite model delay power spectrum. This uncertainty is dictated by uncertainties in spectral indices of foreground models inside the delay horizon delay limits and by thermal noise fluctuations outside}. \label{fig:pitchfork-baselines}}
\end{figure}
